Cubical Types with Positive and Negative Polarities

### 2014-05-19 ring.tex
small typos

%% double-blind reviewing...
\title{Cubical Types}
\authorinfo{}{}{}
\maketitle

@@

Int construction of G construction is neat. As Neel explains, given
first-order types and feedback you get higher-order functions. But if you do
the construction on the additive structure, you lose the multiplicative
structure. It turns out that this is related to a deep open problem in
algebraic topology and homotopy theory that was recently solved. We
``translate'' that solution to a computational type-theoretic world. This has
evident connections to homotopy (type?) theory that remain to be
investigated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Int Construction}

@@

We use $\tau$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\cubt$ for
general $n$-dimensional cubical types. An $n$-dimensional cubical type
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical types of a lower dimension; the 1-dimensional case reduces to plain
first-order types. As we explain below, cubical types are closed under sums,
differences, products, and can express higher-order functions $\cubt_1 \lolli
\cubt_2$ as $(\ominus \cubt_1) \oplus \cubt_2$. The syntax of the types is
summarized below:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \\
\\
\cubt &::=& \tau \alt \nodet{\cubt_1}{\cubt_2} 
      \alt \cubt \oplus \cubt 
      \alt \cubt \otimes \cubt 
      \alt \ominus \cubt 
\end{array}\]
Examples of 1-dimensional cubical types include all the regular first-order
types $\tau$. A 2-dimensional cubical type looks like
$\nodet{\tau_1}{\tau_2}$ which intuitively corresponds to the difference
$\tau_1 - \tau_2$ of the two types. A 3-dimensional cubical type looks like
$\nodet{(\nodet{\tau_1}{\tau_2})}{(\nodet{\tau_3}{\tau_4})}$ which
intuitively corresponds to the iterated difference of types
$(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
encode the sign as follows: a color that is the same as its enclosing color
is positive and a color that differs from its enclosing color is negative. It
is possible for cubical types not be balanced in general but it is always
possible to extend the shorter subtree to an arbitrary depth using the fact
(to be justified later) that~$\tau$ is equivalent to $\nodet{\tau}{0}$.

@@

A few examples might provide some needed intuition. First we note that the
general definitions of sum and product reduce to the conventional sums and
products for 1-dimensional types. Also the sum of 2-dimensional types reduces
to the sum used in the \textbf{Int} construction. The interesting new
development is the product definition. The example in Fig.~\ref{mult} shows
that in general the product of an $n$-dimensional type with an
$m$-dimensional type gives an $n+m$-dimensional type. For low dimensions such
as the ones used in the example above, it is possible to visualize what is
happening in coordinate space.

>>>>

%% double-blind reviewing...
\title{Cubical Types with Positive and Negative Polarities}
\authorinfo{}{}{}
\maketitle

@@

The \textbf{Int} construction or the $\mathcal{G}$ construction are neat. As
Neel K. explains, given first-order types and feedback you get higher-order
functions. But if you do the construction on the additive structure, you lose
the multiplicative structure. It turns out that this is related to a deep
open problem in algebraic topology and homotopy theory that was recently
solved. We ``translate'' that solution to a computational type-theoretic
world. This has evident connections to homotopy (type?) theory that remain to
be investigated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The \textbf{Int} Construction}

@@

We use $\tau$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\cubt$ for
general $n$-dimensional cubical types. An $n$-dimensional cubical type
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical types of a lower dimension; the 1-dimensional case reduces to plain
first-order types. As we explain below, cubical types are closed under sums,
differences, products, and can express higher-order functions $\cubt_1 \lolli
\cubt_2$ as $(\ominus \cubt_1) \oplus \cubt_2$. The syntax of the types is
summarized below:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \\
\\
\cubt &::=& \tau \alt \nodet{\cubt_1}{\cubt_2} 
      \alt \cubt \oplus \cubt 
      \alt \cubt \otimes \cubt 
      \alt \ominus \cubt 
\end{array}\]
Examples of 1-dimensional cubical types include all the regular first-order
types $\tau$. A 2-dimensional cubical type looks like
$\nodet{\tau_1}{\tau_2}$ which intuitively corresponds to the difference
$\tau_1 - \tau_2$ of the two types. A 3-dimensional cubical type looks like
$\nodet{(\nodet{\tau_1}{\tau_2})}{(\nodet{\tau_3}{\tau_4})}$ which
intuitively corresponds to the iterated difference of types
$(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
encode the sign as follows: white in the outermost box corresponds to a
positive polarity and gray in the outermost box corresponds to a negative
polarity; then, as we move inwards the polarities alternate for every change
of color. It is possible for cubical types not be balanced in general but it
is always possible to extend the shorter subtree to an arbitrary depth using
the fact (to be justified later) that~$\tau$ is equivalent to
$\nodet{\tau}{0}$.

@@

A few examples might provide some needed intuition. First we note that the
general definitions of sum and product reduce to the conventional sums and
products for 1-dimensional types. Also the sum of 2-dimensional types reduces
to the sum used in the \textbf{Int} construction. The interesting new
development is the product definition. The example in Fig.~\ref{mult} shows
that in general the product of an $n$-dimensional type with an
$m$-dimensional type gives an $(n+m)$-dimensional type. For low dimensions
such as the ones used in the example above, it is possible to visualize what
is happening in coordinate space (see Fig.~\ref{multv}).
